---
title: "EBC MLS R Demo"
output: html_notebook
author: Eliot McKinley
date: 2018-03-09
---

This R Notebook introduces and gives examples of basic data analysis in R. The example data set is statistics from the 2017 Major League Soccer season derived from www.americansocceranalysis.com and attendance data scraped from Wikipedia.

A few references:
https://www.rstudio.com/wp-content/uploads/2016/10/r-cheat-sheet-3.pdf
https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf
https://www.rstudio.com/wp-content/uploads/2015/08/ggplot2-cheatsheet.pdf
https://www.statmethods.net/stats/index.html
https://data.bits.vib.be/pub/trainingen/RIntro/slides.pdf

Load Data and inspect
```{r}
player.data=read.csv('MLS 2017.csv')
head(player.data)
summary(player.data)

#To access columns in data frame use $
Players=player.data$Player
head(as.character(Players))
```

Getting Frequencies
```{r}
table(player.data$Pos,player.data$Goals)
```

Plotting using base graphics. Good for quick looks at data, but not recommended for publication quality plots.
```{r}
plot(player.data$Goals~player.data$Shots)
plot(player.data$Goals~player.data$Pos)
```

The ggplot2 package allows for much prettier graphs that are more suitable for publication quality figures.
```{r}
library(ggplot2) #load the library functions
ggplot(player.data, aes(x=Shots, y=Goals))+
  geom_point()

ggplot(player.data, aes(x=Pos, y=Goals))+
  geom_boxplot()

```

ggplot2 allows control of almost all aspects of the plot.
```{r}
#load the library functions
library(ggplot2) 

#change the color of the boxplots
ggplot(player.data, aes(x=Pos, y=Goals, fill=Pos))+ 
  geom_boxplot()

#reorder the x axis
ggplot(player.data, aes(x=Pos, y=Goals, color=Pos))+ 
  geom_boxplot()+scale_x_discrete(limits= c('GK','D','M','F')) 

#Bar graph with dots and error bars
ggplot(player.data, aes(x=Pos, y=Goals, fill=Pos))+ 
  geom_bar(stat = "summary", fun.y = "mean")+
  stat_summary(fun.data = mean_se, geom = "errorbar", width=.1)+
  geom_jitter(height=0, width=.3)+
  scale_x_discrete(limits= c('GK','D','M','F')) 

#Violin plot
ggplot(player.data, aes(x=Pos, y=Goals, fill=Pos))+ 
  geom_violin()+
  scale_x_discrete(limits= c('GK','D','M','F')) 

#Prism-like dot plot
ggplot(subset(player.data,Goals>5), aes(x=Pos, y=Goals))+
  geom_dotplot(binaxis = "y", stackdir = "center")+
  stat_summary(fun.y=mean, fun.ymin=mean, fun.ymax=mean, 
                         geom="crossbar", width=0.7)

ggplot(subset(player.data,Goals>5), aes(x=Pos, y=Shots))+
  geom_jitter(width=.1, size=3)+
  stat_summary(fun.y=mean, fun.ymin=mean, fun.ymax=mean, 
                         geom="crossbar", width=0.7)


#histogram
ggplot(player.data, aes(Goals))+
  geom_histogram(breaks=seq(0,25, by = 1))

```

Using ggplot with time series data
```{r}
#Load in time series data
team.attendance=read.csv('MLS.attendance.csv')
names(team.attendance)=gsub('X','',names(team.attendance)) #modify names to get dates
library(reshape2)
team.attendance=melt(team.attendance, by=c('Team','Primary')) #reshape so in tidy format
names(team.attendance)[3:4]=c('Year', 'Attendance')
team.attendance$Year=as.numeric(as.character(team.attendance$Year)) #convert year to number
team.attendance=na.omit(team.attendance)
team.attendance=team.attendance[order(team.attendance$Team),]

#Plot lines of time series data
ggplot(team.attendance, aes(x=Year, y=Attendance))+
  geom_line(aes(group=Team))

#Color the lines
ggplot(team.attendance, aes(x=Year, y=Attendance))+
  geom_line(aes(group=Team, color=Team))+
  geom_point(aes(color=Team), size=.5)+
  scale_color_manual(values=as.character(unique(team.attendance$Primary)))

#Bar plots
ggplot(team.attendance, aes(x=Team, y=Attendance))+
  geom_bar(aes(group=Team, fill=Team), stat = "summary", fun.y = "mean")+
  scale_fill_manual(values=as.character(unique(team.attendance$Primary)))+
  stat_summary(fun.data = mean_se, geom = "errorbar", width=.1)
```

Changing elements of the theme allows for more customization
```{r}  
# Modify X axis Labels
ggplot(team.attendance, aes(x=Team, y=Attendance))+
  geom_bar(aes(group=Team, fill=Team), stat = "summary", fun.y = "mean")+
  scale_fill_manual(values=as.character(unique(team.attendance$Primary)))+
  stat_summary(fun.data = mean_se, geom = "errorbar", width=.1)+
  theme(axis.text.x = (element_text(size=14,angle=45, hjust=1, vjust=1)))

library(ggthemes)

#The Economist
ggplot(team.attendance, aes(x=Team, y=Attendance))+
  geom_bar(aes(group=Team, fill=Team), stat = "summary", fun.y = "mean")+
  scale_fill_manual(values=as.character(unique(team.attendance$Primary)))+
  stat_summary(fun.data = mean_se, geom = "errorbar", width=.1)+
  theme_economist()+
  theme(axis.text.x = (element_text(size=14,angle=45, hjust=1, vjust=1)))
  

#Five Thirty Eight
ggplot(team.attendance, aes(x=Team, y=Attendance))+
  geom_bar(aes(group=Team, fill=Team), stat = "summary", fun.y = "mean")+
  scale_fill_manual(values=as.character(unique(team.attendance$Primary)))+
  stat_summary(fun.data = mean_se, geom = "errorbar", width=.1)+
  theme_fivethirtyeight()+
  theme(axis.text.x = (element_text(size=14,angle=45, hjust=1, vjust=1)))

#Excel
ggplot(team.attendance, aes(x=Team, y=Attendance))+
  geom_bar(aes(group=Team, fill=Team), stat = "summary", fun.y = "mean")+
  scale_fill_manual(values=as.character(unique(team.attendance$Primary)))+
  stat_summary(fun.data = mean_se, geom = "errorbar", width=.1)+
  theme_excel()+
  theme(axis.text.x = (element_text(size=14,angle=45, hjust=1, vjust=1)))

#Minimal
ggplot(team.attendance, aes(x=Team, y=Attendance))+
  geom_bar(aes(group=Team, fill=Team), stat = "summary", fun.y = "mean")+
  scale_fill_manual(values=as.character(unique(team.attendance$Primary)))+
  stat_summary(fun.data = mean_se, geom = "errorbar", width=.1)+
  theme_minimal()+
  theme(axis.text.x = (element_text(size=14,angle=45, hjust=1, vjust=1)))+
  labs(title='Attendance: 1996-2017', y='Announced Attendance', x='Franchise')



#Minimal- plus reorder based upon attendance

team.attendance$Team=factor(team.attendance$Team, levels=levels(reorder(team.attendance$Team, -team.attendance$Attendance,mean)))
ggplot(team.attendance, aes(x=Team, y=Attendance))+
  geom_bar(aes(group=Team, fill=Team), stat = "summary", fun.y = "mean")+
  scale_fill_manual(values=levels(reorder(team.attendance$Primary,-team.attendance$Attendance, mean)))+
  stat_summary(fun.data = mean_se, geom = "errorbar", width=.1)+
  theme_minimal()+
  theme(axis.text.x = (element_text(size=14,angle=45, hjust=1, vjust=1)))+
  labs(title='Attendance: 1996-2017', y='Announced Attendance', x='Franchise')+
  theme(legend.position = "none")
```

Basic Statistical Tests
Check out the ggpubr package: http://www.sthda.com/english/rpkgs/ggpubr/
```{r}
Teams=subset(team.attendance, Team %in% c('CLB','CHI'))
ggplot(Teams, aes(x=Team, y=Attendance))+geom_point()+stat_summary(fun.data = mean_se, geom = "crossbar", width=.1, color="red")

t.test(Teams$Attendance~Teams$Team)
wilcox.test(Teams$Attendance~Teams$Team)

library(ggpubr)
ggplot(Teams, aes(x=Team, y=Attendance))+geom_point()+
  stat_summary(fun.data = mean_se, geom = "crossbar", width=.1, color="red")+
  stat_compare_means()+
  scale_y_continuous(limits=c(10000,20000))

Teams=subset(team.attendance, Team %in% c('CLB','CHI') & Year > 1998)
wilcox.test(Teams$Attendance~Teams$Team, paired=TRUE)
ggpaired(Teams, x = "Team", y = "Attendance",
                color = "Team", palette ='jco',
                add = "jitter", shape = "Team",
         line.color = "gray")+
  stat_compare_means(comparisons=list(c('CLB','CHI')), paired=TRUE)


Teams=subset(team.attendance, Team %in% c('CLB','COL','DCU','FCD','LAG','NER','NYRB','SJE','SKC'))
ggplot(Teams, aes(x=Team, y=Attendance))+
  geom_jitter(width=.1)+
  stat_summary(fun.data = mean_se, geom = "crossbar", width=.2, color="red")+
  stat_compare_means(label = "p.signif", comparisons=list(c('NYRB','COL'),c('NER','DCU'),c('SKC','CLB')))+
  stat_compare_means(method='anova', label.y=10000)+
  theme_minimal()


ggboxplot(Teams, x='Team', y='Attendance', color='Team', add='jitter', legend='none')+
  geom_hline(yintercept = mean(Teams$Attendance), linetype = 2)+
  stat_compare_means(method = "anova", label.y = 10000)+
  stat_compare_means(label = "p.signif", method = "wilcox.test",
                     ref.group = ".all.", label.y=30000) 

ggboxplot(Teams, x='Team', y='Attendance', color='Team', add='jitter', legend='none')+
  geom_hline(yintercept = mean(Teams$Attendance), linetype = 2)+
  stat_compare_means(method = "anova", label.y = 10000)+
  stat_compare_means(label = "p.signif", method = "wilcox.test",
                     ref.group = "CLB", hide.ns=TRUE, label.y=30000) 

anov=aov(Teams$Attendance~Teams$Team)

summary(anov)
TukeyHSD(anov)

```
Correlations
```{r}
ggplot(Teams, aes(x=Year, y=Attendance))+geom_point()

cor.test(Teams$Attendance,Teams$Year, use="complete.obs")

library(ggpmisc)
formula = y ~ x

ggplot(Teams, aes(x=Year, y=Attendance))+geom_point(aes(color=Team))+
  scale_color_manual(values=as.character(unique(team.attendance$Primary)))+
  geom_smooth(method = "lm", se=FALSE)+
  stat_poly_eq(aes(label = paste(..eq.label..,..rr.label.., sep = "~~~")), 
               label.x.npc = "left", label.y.npc = 0.9,
               formula = formula, parse = TRUE, size = 5)+
  theme_minimal()
```

Principal Components Analysis see: https://www.r-bloggers.com/principal-component-analysis-in-r/
```{r}
#clean up data for PCA
PCA.data=na.omit(player.data)
player.PCA=prcomp(PCA.data[,-c(1:3,32:34)])
PCA.data=data.frame(PCA.data,player.PCA$x)

ggplot(PCA.data, aes(x=PC1, y=PC2, color=Pos))+
  geom_point()+
  theme_minimal()

```

tSNE see: https://distill.pub/2016/misread-tsne/
```{r}

library(Rtsne)

tSNE.data=na.omit(player.data)
set.seed(42)
tSNE=Rtsne(as.matrix(tSNE.data[,c(4:31,35)]), perplexity=5,pca_center = TRUE, pca_scale = TRUE, verbose = TRUE)

tSNE.data$tSNE1=tSNE$Y[,1]
tSNE.data$tSNE2=tSNE$Y[,2]

ggplot(tSNE.data, aes(x=tSNE1, y=tSNE2, color=Pos))+
  geom_point()+
  theme_minimal()

#Uncomment these lines to install the MxIF package
#library(devtools)
#install_github('etmckinley/MxIF')
library(MxIF)

MxIF.tsne.plot(tSNE.data,8)+labs(title='Goals')

```



Hierarchical Clustering see: https://www.stat.berkeley.edu/~s133/Cluster2a.html
```{r}
clust.data=na.omit(player.data) #remove NAs in data
clust.data=subset(clust.data, Min>2300)
hclust.data=clust.data[,-c(1:3,32:34)] #remove non numeric variables
rownames(hclust.data)=clust.data$Player

#standardize the data
medians=apply(hclust.data,2,median) #apply the median function to each column (2), or doing rows, apply(clust.data,1,median)
mads=apply(hclust.data,2,mad)
hclust.data=scale(hclust.data, center=medians, scale=mads)

#create distance matrix
hclust.dist=dist(hclust.data)

hclust.data=hclust(hclust.dist)

plot(hclust.data, cex=0.5)

library(ggdendro)

p=ggdendrogram(hclust.data, rotate=TRUE, size=0.3)+labs(title='Hierachical Clustering, Min>2300')
p

p=ggsave('player_dendro.png', plot=p, height=12, width=5)

```
